A Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-02-08 02-ceu.tex
Many changes:  - \code definition (problems with the font)  - "every" as a primitive in the concrete and formal language  - removed "Concrete Language Mapping" section (no space), included the most relevant parts in the semantics discussion

In \CEU, every execution path within loops must contain at least one
\code{await} statement~\cite{ceu.sensys13,esterel.primer}.
%
This restriction, which is statically checked by the compiler, ensures that
every reaction runs in bounded time, eventually terminating with all trails
blocked in \code{await} statements.
%
\CEU has an additional restriction, which it shares with Esterel and
synchronous languages in general~\cite{esterel.preemption}: computations that
take a non-negligible time to run (e.g., cryptography or image processing
algorithms) violate the zero-delay hypothesis, and thus cannot be directly
implemented.

@@

Listing~\ref{lst.ceu} shows a complete example in \CEU that blinks a LED
whenever a radio packet is received, terminating with a button press always
with the LED off.
%
The implementation first declares the \code{BUTTON} and \code{RADIO_RECV} as
input events (ln.~1--2).
Then, it uses a \code{par/or} composition to run two activities in parallel:
a single statement that waits for a button press before terminating (ln.~4),
and an endless loop that blinks the LED on and off (ln.~9--14).
The \code{finalize} clause (ln.~6--8) ensures that, no matter how its enclosing
trail terminates, the LED will be unconditionally turned off (ln.~7).

@@

In \CEU, any identifier prefixed with an underscore (e.g., \code{_led}) is
passed unchanged to the underlying~C compiler.
%
Therefore, access to~C is straightforward and syntactically traceable.
%
To ensure that programs operate under the synchronous hypothesis, the compiler
environment should only provide access to~C operations that can be assumed to
be instantaneous, such as non-blocking~I/O and simple accesses to data
structures.

@@

At first sight, an event iterator for event \code{A} seems like a syntactic
sugar for a \code{loop} followed by and \code{await A}.
However, event iterators have an important restriction to enforce reaction
termination, as we discuss in Section~\ref{sec.sem}:
    they cannot contain \code{break} statements.

@@

\CEU also enforces the use of \code{finalize} for system calls that deal with
pointers representing resources, as illustrated in the two examples of
Figure~\ref{lst.fin.ceu}:
%
\begin{itemize}
    \item If \CEU \textbf{passes} a pointer to a system call (ln. \ax:5), the
          pointer represents a \textbf{local} resource (ln. \ax:2) that
          requires finalization (ln. \ax:7).
    \item If \CEU \textbf{receives} a pointer from a system call return
          (ln.  \bx:4), the pointer represents an \textbf{external} resource
          (ln. \bx:2) that requires finalization (ln. \bx:6).
\end{itemize}
%
\CEU tracks the interaction of system calls with pointers and requires
finalization clauses to accompany them.
%
In the example in Figure~\ref{lst.fin.ceu}.a, the local variable \code{msg}
(ln. 2) is an internal resource passed as a pointer to \code{_send} (ln. 5),
which is an asynchronous call that transmits the buffer in the background.
If the block aborts (ln. 11) before receiving an acknowledge from the
environment (ln. 9), the local \code{msg} goes out of scope and the external
transmission now holds a \emph{dangling pointer}.
The finalization ensures that the transmission also aborts (ln. 7).
%
In the example in Figure~\ref{lst.fin.ceu}.b, the call to \code{_fopen} (ln.
4) returns an external file resource as a pointer.
If the block aborts (ln. 12) during the \code{await A} (ln. 9), the file
remains open as a \emph{memory leak}.
The finalization ensures that the file closes properly (ln. 6).
%
In both cases, the code does not compile without the \code{finalize}
construct.%
\footnote{
The compiler only forces the programmer to write the finalization clause, but
cannot check if it actually handles the resource properly.
}

>>>>

In \CEU, every execution path within loops must contain at least one
\code{await} statement to an external input
event~\cite{ceu.sensys13,esterel.primer}.
% TODO: or every
%
This restriction, which is statically checked by the compiler, ensures that
every reaction runs in bounded time, eventually terminating with all trails
blocked in \code{await} statements.
%
\CEU has an additional restriction, which it shares with Esterel and
synchronous languages in general~\cite{esterel.preemption}: computations that
take a non-negligible time to run (e.g., cryptography or image processing
algorithms) violate the zero-delay hypothesis, and thus cannot be directly
implemented.

@@

Listing~\ref{lst.ceu} shows a complete example in \CEU that blinks a LED
whenever a radio packet is received, terminating with a button press always
with the LED off.
%
The implementation first declares the \code{BUTTON} and \code{RADIO\_RECV} as
input events (ln.~1--2).
Then, it uses a \code{par/or} composition to run two activities in parallel:
a single statement that waits for a button press before terminating (ln.~4),
and an endless loop that blinks the LED on and off (ln.~9--14).
The \code{finalize} clause (ln.~6--8) ensures that, no matter how its enclosing
trail terminates, the LED will be unconditionally turned off (ln.~7).

@@

In \CEU, any identifier prefixed with an underscore (e.g., \code{\_led}) is
passed unchanged to the underlying~C compiler.
%
Therefore, access to~C is straightforward and syntactically traceable.
%
To ensure that programs operate under the synchronous hypothesis, the compiler
environment should only provide access to~C operations that can be assumed to
be instantaneous, such as non-blocking~I/O and simple accesses to data
structures.

@@

At first sight, event iteration such as ``\code{every e do <...> end}'' seems
to be equivalent to ``\code{loop do await e; <...> end}''.
However, the \code{loop} variation would not compile because it does not
contain a path to an external input \code{await} (\code{e} is an internal
event).
Event iterators have syntactic restrictions and cannot contain \code{await} or
\code{break} statements.
The absence of \code{break} guarantees that an iterator never terminates from
itself, essentially behaving as safe blocking point in the program.

@@

\CEU also enforces the use of \code{finalize} for system calls that deal with
pointers representing resources, as illustrated in the two examples of
Figure~\ref{lst.fin.ceu}:
%
\begin{itemize}
    \item If \CEU \textbf{passes} a pointer to a system call (ln. \ax:5), the
          pointer represents a \textbf{local} resource (ln. \ax:2) that
          requires finalization (ln. \ax:7).
    \item If \CEU \textbf{receives} a pointer from a system call return
          (ln.  \bx:4), the pointer represents an \textbf{external} resource
          (ln. \bx:2) that requires finalization (ln. \bx:6).
\end{itemize}
%
\CEU tracks the interaction of system calls with pointers and requires
finalization clauses to accompany them.
%
In the example in Figure~\ref{lst.fin.ceu}.a, the local variable \code{msg}
(ln. 2) is an internal resource passed as a pointer to \code{\_send} (ln. 5),
which is an asynchronous call that transmits the buffer in the background.
If the block aborts (ln. 11) before receiving an acknowledge from the
environment (ln. 9), the local \code{msg} goes out of scope and the external
transmission now holds a \emph{dangling pointer}.
The finalization ensures that the transmission also aborts (ln. 7).
%
In the example in Figure~\ref{lst.fin.ceu}.b, the call to \code{\_fopen} (ln.
4) returns an external file resource as a pointer.
If the block aborts (ln. 12) during the \code{await A} (ln. 9), the file
remains open as a \emph{memory leak}.
The finalization ensures that the file closes properly (ln. 6).
%
In both cases, the code does not compile without the \code{finalize}
construct.%
\footnote{
The compiler only forces the programmer to write the finalization clause, but
cannot check if it actually handles the resource properly.
}

### 2018-02-08 03-semantics.tex
Many changes:  - \code definition (problems with the font)  - "every" as a primitive in the concrete and formal language  - removed "Concrete Language Mapping" section (no space), included the most relevant parts in the semantics discussion

\begin{lstlisting}[
  %numbers=left,
  basicstyle=\ttfamily\footnotesize,
  float=h,
  caption={Reduced syntax of \CEU.},
  label={lst.formal.syntax},
  mathescape=true
]
                                   // primary expressions
  p ::= mem(id)                    (any memory access to `id')
      $|$ awaitExt(id)               (await external event `id')
      $|$ awaitInt(id)               (await internal event `id')
      $|$ emitInt(id)                (emit internal event `id')
      $|$ break                      (loop escape)
                                   // compound expressions
      $|$ if mem(id) then p else p   (conditional)
      $|$ p ; p                      (sequence)
      $|$ loop p                     (repetition)
      $|$ p and p                    (par/and)
      $|$ p or p                     (par/or)
      $|$ fin p                      (finalization)
                                   // derived by semantic rules
      $|$ p @loop p                  (unwinded loop)
      $|$ p @and q                   (unwinded par/and)
      $|$ p @or q                    (unwinded par/or)
      $|$ @canrun(n)                 (can run on stack level `n')
      $|$ @nop                       (terminated expression)
\end{lstlisting}

Listing~\ref{lst.formal.syntax} shows the syntax for a subset of \CEU that is 
sufficient to describe all semantic peculiarities of the language.
%
Except for $fin$ and the expressions used internally by the semantics, which
are prefixed with a character $@$, all other expressions are equivalent to
their counterparts in the concrete language.

The $mem(id)$ primitive represents all accesses, assignments, system calls, and
output events that affect a memory location identified by $id$.
%
According to the synchronous hypothesis of \CEU, $mem$ expressions are 
considered to be atomic and instantaneous.
%
As the challenging parts of \CEU reside on its control structures, we are not 
concerned here with a precise semantics for side effects, but only with their 
occurrences in programs.
%
%The special notation $nop$ is used to represent an innocuous $mem$ expression 
%(it can be thought as a synonym for $mem(\epsilon)$, where $\epsilon$ is an 
%unused identifier).
%
Note that $mem$, $awaitExt$, and $awaitInt$ / $emitInt$ expressions do not
share identifiers, i.e., an identifier is either a variable, an external event,
or an internal event.

@@

The $\NST$ transition rules for primary expressions are as follows:
%
{ \setlength{\jot}{20pt}
\begin{align*}
\LL mem(id), n, \epsilon \RR &\NST
\LL @nop, n, \epsilon \RR
    & \textbf{(mem)}        \\
%%%
\LL emit(id), n, \epsilon \RR &\NST
\LL @canrun(n), n, id \RR
    & \textbf{(emitInt)}    \\
%%%
\LL @canrun(n), n, \epsilon \RR &\NST
\LL @nop, n, \epsilon \RR
    & \textbf{(canrun)}     \\
\end{align*}
}
%
A $mem$ operation becomes a $@nop$ which indicates termination (rule
\textbf{mem}).
An $emitInt(id)$ generates an event $id$ and transits to a $@canrun(n)$ which
can only resume at level $n$ (rule \textbf{emitInt}).
Since nested rules cannot transit with $e \neq \epsilon$, this rule will cause
rule \textbf{push} to execute at the outer level, creating a new level $n+1$ on
the stack.
Also, with the new stack level, the resulting $@canrun(n)$ itself cannot
transit, providing the desired stack-based semantics for internal events.

@@

For a parallel $@and$, if one of the sides terminates, the composition is
simply substituted by the other side (rules \textbf{and-nop1} and
\textbf{and-nop2}.
%
For a parallel $or$, if one of the sides terminates, the whole composition 
terminates, also applying the $clear$ function to properly finalize the aborted 
side (rules \textbf{or-nop1} and \textbf{or-nop2}):
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \LL (@nop~@and~q), n, \epsilon \RR \NST \LL q,n,\epsilon \RR
    & \textbf{(and-nop1)}   \\
%%%
& \LL (p~@and~@nop), n, \epsilon \RR \NST \LL p,n,\epsilon \RR
    & \textbf{(and-nop2)}   \\
%%%
& \LL (@nop~@or~q), n, \epsilon \RR \NST \LL clear(q),n,\epsilon \RR
    & \textbf{(or-nop1)}   \\
%%%
& \frac
    { \DS isblocked(n,p) }
%   -----------------------------------------------------------
    { \DS \LL (p~@or~@nop), n, \epsilon \RR \NST \LL clear(p),n,\epsilon \RR }
    & \textbf{(or-nop2)}   %\\
\end{eqnarray*}
}

%
The $clear$ function (Figure~\ref{fig.formal.clear}) concatenates all active
$fin$ bodies of the side being aborted, so that they execute before the
composition rejoins.
Note that there are no transition rules for $fin$ expressions.
This is because once reached, a $fin$ expression halts and will only execute 
when it is aborted by a trail in parallel and is expanded by the $clear$ 
function.
%In Section~\ref{sec.formal.fins}, we show how to map a finalization block in 
%the concrete language to a $fin$ in the formal semantics.
%
Note also that there is a syntactic restriction that $fin$ bodies cannot
contain awaiting expressions ($awaitExt$, $awaitInt$, and $fin$), i.e., they
are guaranteed to execute entirely within a reaction.

Finally, a $break$ in one of the sides in parallel escapes the closest 
enclosing $loop$, properly aborting the other side by applying the $clear$ 
function:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \LL (break~@and~q), n, \epsilon \RR \NST \LL (clear(q)~;~break),n,\epsilon \RR
    & \textbf{(and-brk1)}   \\
%%%
& \frac
    { \DS isblocked(n,p) }
%   -----------------------------------------------------------
    { \DS \LL (p~@and~break), n, \epsilon \RR \NST \LL (clear(p)~;~break),n,\epsilon \RR }
    & \textbf{(and-brk2)}   \\
%%%
& \LL (break~@or~q),n,\epsilon \RR \NST \LL (clear(q)~;~break),n,\epsilon \RR
    & \textbf{(or-brk1)}   \\
%%%
& \frac
    { \DS isblocked(n,p) }
%   -----------------------------------------------------------
    { \DS \LL (p~@or~break),n,\epsilon \RR \NST \LL (clear(p)~;~break),n,\epsilon \RR }
    & \textbf{(or-brk2)}   %\\
\end{eqnarray*}
}
%
A reaction eventually blocks in $awaitExt$, $awaitInt$, $fin$, and $@canrun$
expressions in parallel trails.
%
If no trails are blocked in $@canrun$ expressions, it means that the program 
cannot advance in the current reaction.
%
However, $@canrun$ expressions can still resume in lower stack indexes and will
eventually resume in the current reaction (see rule \textbf{pop}).

\begin{figure}
{\small
\begin{align*}
  bcast(e, awaitExt(e)) &= @nop                         \\
  bcast(e, awaitInt(e)) &= @nop                         \\
  bcast(e, @canrun(n))  &= @canrun(n)                   \\
  bcast(e, fin~p)       &= fin~p                        \\
  bcast(e, p~;~q)       &= bcast(e,p)~;~q               \\
  bcast(e, p~@loop~q)   &= bcast(e,p)~@loop~q           \\
  bcast(e, p~@and~q)    &= bcast(e,p)~@and~bcast(e,q)   \\
  bcast(e, p~@or~q)     &= bcast(e,p)~@or~bcast(e,q)    \\
  bcast(e, \_)          &= \bot \2 (mem,emitInt,break,if,  \\
                                 & \5\5 loop,and,or,@nop) %\\
\end{align*}
}
\caption{
The function $bcast$ awakes awaiting trails matching the event by converting
\code{awaitExt} and \code{awaitInt} to \code{@nop} expressions.
\label{fig.bcast}
}
\end{figure}

\begin{figure}
{\small
\begin{align*}
  isblocked(n, \1 awaitExt(id)) &= true                                   \\
  isblocked(n, \1 awaitInt(id)) &= true                                   \\
  isblocked(n, \1 @canrun(m))   &= (n > m)                                \\
  isblocked(n, \1 fin~p)        &= true                                   \\
  isblocked(n, \1 p~;~q)        &= isblocked(n,p)                         \\
  isblocked(n, \1 p~@loop~q)    &= isblocked(n,p)                         \\
  isblocked(n, \1 p~@and~q)     &= isblocked(n,p) \wedge isblocked(n,q)   \\
  isblocked(n, \1 p~@or~q)      &= isblocked(n,p) \wedge isblocked(n,q)   \\
  isblocked(n, \1 \_)           &= false \2 (mem,emitInt,break,if,        \\
                                & \5\5\5\1 loop,and,or,@nop)   %\\
\end{align*}
}
\caption{
The predicate $isblocked$ is true only if all branches in parallel are blocked
waiting for events, finalization clauses, or certain stack levels.
\label{fig.isblocked}
}
\end{figure}

\begin{figure}[b]
{\small
\begin{align*}
  clear( awaitExt(e) ) &= @nop                  \\
  clear( awaitInt(e) ) &= @nop                  \\
  clear( @canrun(n) )  &= @nop                  \\
  clear( fin~p )       &= p                     \\
  clear( p~;~q )       &= clear(p)              \\
  clear( p~@loop~q )   &= clear(p)              \\
  clear( p~@and~q )    &= clear(p)~;~clear(q)   \\
  clear( p~@or~q )     &= clear(p)~;~clear(q)   \\
  clear( \_ )          &= \bot \2 (mem,emitInt,break,if, \\
                                  & \5\5 loop,and,or,@nop) %\\ 
\end{align*}
}
\caption{
The function $clear$ extracts $fin$ expressions in parallel and put their 
bodies in sequence.
\label{fig.formal.clear}
}
\end{figure}

\subsection{Concrete Language Mapping}

Most statements from \CEU (``concrete \CEU'') map directly to those presented 
in the reduced syntax in Figure~\ref{lst.formal.syntax} (``abstract \CEU'').
%, namely, the \code{if}, \code{`;'}, \code{loop}, \code{par/and}, and 
%\code{par/or}.
For instance, the \code{if} in the concrete language behaves exactly like the 
$if$ in the formal semantics.
However, there are some significant mismatches between the concrete and 
abstract \CEU, and we (informally) present appropriate mappings in this 
section.
%
Again, we are not considering side-effects, which are all mapped to the $mem$ 
semantic construct.

\begin{comment}
$
    \code{if}      \mapsto if   ,\2
    \code{';'}     \mapsto ~';' ,\2
    \code{loop}    \mapsto loop ,\2
    \code{par/and} \mapsto and  ,\2
    \code{par/or}  \mapsto or
$.
\end{comment}

\subsubsection{await and emit}

The concrete \code{await} and \code{emit} primitives support communication of 
values between them.
In the two-step translation in Figure~\ref{lst.map.emit.await}, we start with 
the concrete program in \CEU (a), which communicates the value $1$ between the 
\code{emit} and \code{await} in parallel.
In the intermediate translation (b), we include the shared variable \code{e\_} 
to hold the value being communicated between the two trails in order to 
simplify the \code{emit}.
Finally, we convert the program into the equivalent in the abstract syntax (c), 
translating side-effect statements into $mem$ expressions.
External events have a similar translation, i.e., each external event requires 
a corresponding variable that is explicitly set by the environment before each 
reaction.

\begin{figure}[t]
\begin{minipage}[t]{0.32\linewidth}
\begin{lstlisting}
par/or do
  <...>
  emit e => 1;
with
  v = await e;
  _printf("%d\n",v);
end

.
\end{lstlisting}
\centering\small{(a)}
\end{minipage}
%
\begin{minipage}[t]{0.32\linewidth}
\begin{lstlisting}
par/or do
  <...>
  e_ = 1;
  emit e;
with
  await e;
  v = e_;
  _printf("%d\n",v);
end
\end{lstlisting}
\centering\small{(b)}
\end{minipage}
%
\begin{minipage}[t]{0.32\linewidth}
\begin{lstlisting}
<...> ; mem ; emit(e)
        or
await(e) ; mem ; mem

.
\end{lstlisting}
\centering\small{(c)}
\end{minipage}
\caption{
Two-step translation from concrete to abstract \code{emit} and \code{await} 
expressions.
The concrete code in (a) communicates the value \code{1} from the \code{emit} 
to the \code{await}.
The abstract code in (c) uses a shared variable to hold the value.
\label{lst.map.emit.await}
}
\end{figure}

\subsubsection{Finalization Blocks}
\label{sec.formal.fins}

The biggest mismatch between concrete and abstract \CEU is regarding the 
\code{finalize} blocks, which require more complex modifications in the program
for a proper mapping using $fin$ expressions.
In the three-step translation in Figure~\ref{lst.map.fin}, we start with a 
concrete program (a) that uses a \code{finalize} to safely \code{\_release} the 
reference to \code{ptr} kept after the call to \code{\_hold}.
In the translation, we first need to catch the outermost \code{do-end} 
termination to run the finalization code.
For this, we translate the block into a \code{par/or} (b) with the original 
body in parallel with a $fin$ expression to run the finalization code.
Note that the $fin$ has no transition rules in the semantics, keeping the 
\code{par/or} alive.
This way, the $fin$ body only executes when the \code{par/or} terminates either 
normally (after the \code{await B}), or aborted from an outer composition.
%
However, the $fin$ still (incorrectly) executes even if the call to 
\code{\_hold} is not reached in the body due to an abort before awaking from 
the \code{await A}.
%
To deal with this issue, for each $fin$ we need a corresponding flag to keep 
track of code that needs to be finalized (c).
%
The flag is initially set to false, avoiding the finalization code to execute.
Only after the call to \code{\_hold} that we set the flag to true and enable 
the $fin$ body to execute.
%
The complete translation substitutes the side-effect operations with $mem$ 
expressions (d).

\begin{figure}
\begin{minipage}[t]{0.26\linewidth}
\begin{lstlisting}
do
  var int* ptr = <...>;
  await A;
  finalize
    _hold(ptr);
  with
    _release(ptr);
  end
  await B;
end

.
\end{lstlisting}
\centering\small{(a)}
\end{minipage}
%
\begin{minipage}[t]{0.26\linewidth}
\begin{lstlisting}
par/or do
  var int* ptr = <...>;
  await A;
  _hold(ptr);
  await B;
with
  { fin
      _release(ptr); }
end

.
\end{lstlisting}
\centering\small{(b)}
\end{minipage}
%
\begin{minipage}[t]{0.26\linewidth}
\begin{lstlisting}
f_ = 0;
par/or do
  var int* ptr = <...>;
  await A;
  _hold(ptr);
  f_ = 1;
  await B;
with
  { fin
      if f_ then
        _release(ptr);
      end }
end
.
\end{lstlisting}
\centering\small{(c)}
\end{minipage}
%
\begin{minipage}[t]{0.18\linewidth}
\begin{lstlisting}
mem;
(
   mem;
   await(A);
   mem;
   mem;
   await(B);
or
   fin
     if mem then
       mem
     else
       nop
)
\end{lstlisting}
\centering\small{(d)}
\end{minipage}
%
\caption{
Three-step translation from concrete to abstract finalization.
\label{lst.map.fin}
}
\end{figure}

>>>>

\begin{lstlisting}[
  %numbers=left,
  basicstyle=\ttfamily\footnotesize,
  float=h,
  caption={Reduced syntax of \CEU.},
  label={lst.formal.syntax},
  mathescape=true
]
                                   // primary expressions
  p ::= mem(id)                    (any memory access to `id')
      $|$ awaitExt(id)               (await external event `id')
      $|$ awaitInt(id)               (await internal event `id')
      $|$ emitInt(id)                (emit internal event `id')
      $|$ break                      (loop escape)
                                   // compound expressions
      $|$ if mem(id) then p else p   (conditional)
      $|$ p ; p                      (sequence)
      $|$ loop p                     (repetition)
      $|$ every id p                 (event iteration)
      $|$ p and p                    (par/and)
      $|$ p or p                     (par/or)
      $|$ fin p                      (finalization)
                                   // derived by semantic rules
      $|$ p @loop p                  (unwinded loop)
      $|$ p @and q                   (unwinded par/and)
      $|$ p @or q                    (unwinded par/or)
      $|$ @canrun(n)                 (can run on stack level `n')
      $|$ @nop                       (terminated expression)
\end{lstlisting}

Listing~\ref{lst.formal.syntax} shows the syntax for a subset of \CEU that is 
sufficient to describe all semantic peculiarities of the language.
%
The $mem(id)$ primitive represents all accesses, assignments, system calls, and
output events that affect a memory location identified by $id$.
%
According to the synchronous hypothesis of \CEU, $mem$ expressions are 
considered to be atomic and instantaneous.
%
As the challenging parts of \CEU reside on its control structures, we are not 
concerned here with a precise semantics for side effects, but only with their 
occurrences in programs.
%
%The special notation $nop$ is used to represent an innocuous $mem$ expression 
%(it can be thought as a synonym for $mem(\epsilon)$, where $\epsilon$ is an 
%unused identifier).
%
Note that $mem$, $awaitExt$, and $awaitInt$ / $emitInt$ expressions do not
share identifiers, i.e., an identifier is either a variable, an external event,
or an internal event.

In the concrete language, the \code{await} and \code{emit} primitives support
communication of values between them, e.g., an \code{emit a(10)} awakes
\code{v=await a} setting \code{v=10}.
To reproduce this functionality in the formal semantics, we can use a shared
variable to hold the value of an \code{emit} and access it after the
corresponding \code{await}.

A ``\code{finalize <A> with <B> end ; <C>}'' in the concrete language is
equivalent to ``\code{<A> ; ((fin <B>) or <C>)}'' in the abstract syntax.
In the concrete language, \code{<A>} and \code{<C>} execute in sequence, and
the finalization code \code{<B>} is implicitly suspended waiting for \code{<C>}
termination.
In the formal language, \code{fin <B>} suspends forever when reached (it is an
awaiting expression that never awakes).
Hence, we need an explicit \code{or} to execute \code{<C>} in parallel, whose
termination aborts \code{fin <B>}, which finally causes \code{<B>} to execute
(by the semantic rules to be discussed).

@@

The $\NST$ transition rules for primary expressions are as follows:
%
{ \setlength{\jot}{20pt}
\begin{align*}
\LL mem(id), n, \epsilon \RR &\NST
\LL @nop, n, \epsilon \RR
    & \textbf{(mem)}        \\
%%%
\LL emit(id), n, \epsilon \RR &\NST
\LL @canrun(n), n, id \RR
    & \textbf{(emitInt)}    \\
%%%
\LL @canrun(n), n, \epsilon \RR &\NST
\LL @nop, n, \epsilon \RR
    & \textbf{(canrun)}     \\
\end{align*}
}
%
A $mem$ operation becomes a $@nop$ which indicates termination (rule
\textbf{mem}).
An $emitInt(id)$ generates an event $id$ and transits to a $@canrun(n)$ which
can only resume at level $n$ (rule \textbf{emitInt}).
Since nested rules can only transit with $e=\epsilon$, this rule causes rule
\textbf{push} to execute at the outer level, creating a new level $n+1$ on
the stack.
Also, with the new stack level, the resulting $@canrun(n)$ itself cannot
transit, providing the desired stack-based semantics for internal events.

@@

For a parallel $@and$, if one of the sides terminates, the composition is
simply substituted by the other side (rules \textbf{and-nop1} and
\textbf{and-nop2}.
%
For a parallel $or$, if one of the sides terminates, the whole composition 
terminates, also applying the $clear$ function to properly finalize the aborted 
side (rules \textbf{or-nop1} and \textbf{or-nop2}):
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \LL (@nop~@and~q), n, \epsilon \RR \NST \LL q,n,\epsilon \RR
    & \textbf{(and-nop1)}   \\
%%%
& \LL (p~@and~@nop), n, \epsilon \RR \NST \LL p,n,\epsilon \RR
    & \textbf{(and-nop2)}   \\
%%%
& \LL (@nop~@or~q), n, \epsilon \RR \NST \LL clear(q),n,\epsilon \RR
    & \textbf{(or-nop1)}   \\
%%%
& \frac
    { \DS isblocked(n,p) }
%   -----------------------------------------------------------
    { \DS \LL (p~@or~@nop), n, \epsilon \RR \NST \LL clear(p),n,\epsilon \RR }
    & \textbf{(or-nop2)}   %\\
\end{eqnarray*}
}
%
The $clear$ function (Figure~\ref{fig.formal.clear}) concatenates all active
$fin$ bodies of the side being aborted, so that they execute before the
composition rejoins.
Note that there are no transition rules for $fin$ expressions.
This is because once reached, a $fin$ expression halts and will only execute 
when it is aborted by a trail in parallel and is expanded by the $clear$ 
function.
%In Section~\ref{sec.formal.fins}, we show how to map a finalization block in 
%the concrete language to a $fin$ in the formal semantics.
%
Note also that there is a syntactic restriction that $fin$ bodies cannot
contain awaiting expressions ($awaitExt$, $awaitInt$, $every$, and $fin$),
i.e., they are guaranteed to execute entirely within a reaction.

Finally, a $break$ in one of the sides in parallel escapes the closest 
enclosing $loop$, properly aborting the other side by applying the $clear$ 
function:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \LL (break~@and~q), n, \epsilon \RR \NST \LL (clear(q)~;~break),n,\epsilon \RR
    & \textbf{(and-brk1)}   \\
%%%
& \frac
    { \DS isblocked(n,p) }
%   -----------------------------------------------------------
    { \DS \LL (p~@and~break), n, \epsilon \RR \NST \LL (clear(p)~;~break),n,\epsilon \RR }
    & \textbf{(and-brk2)}   \\
%%%
& \LL (break~@or~q),n,\epsilon \RR \NST \LL (clear(q)~;~break),n,\epsilon \RR
    & \textbf{(or-brk1)}   \\
%%%
& \frac
    { \DS isblocked(n,p) }
%   -----------------------------------------------------------
    { \DS \LL (p~@or~break),n,\epsilon \RR \NST \LL (clear(p)~;~break),n,\epsilon \RR }
    & \textbf{(or-brk2)}   %\\
\end{eqnarray*}
}
%
A reaction eventually blocks in $awaitExt$, $awaitInt$, $every$, $fin$, and
$@canrun$ expressions in parallel trails.
%
If no trails are blocked in $@canrun$ expressions, it means that the program 
cannot advance in the current reaction.
%
However, $@canrun$ expressions can still resume in lower stack indexes and will
eventually resume in the current reaction (see rule \textbf{pop}).

\begin{figure}
{\small
\begin{align*}
  bcast(e, awaitExt(e)) &= @nop                         \\
  bcast(e, awaitInt(e)) &= @nop                         \\
  bcast(e, every~e~p)   &= p; every e p                 \\
  bcast(e, @canrun(n))  &= @canrun(n)                   \\
  bcast(e, fin~p)       &= fin~p                        \\
  bcast(e, p~;~q)       &= bcast(e,p)~;~q               \\
  bcast(e, p~@loop~q)   &= bcast(e,p)~@loop~q           \\
  bcast(e, p~@and~q)    &= bcast(e,p)~@and~bcast(e,q)   \\
  bcast(e, p~@or~q)     &= bcast(e,p)~@or~bcast(e,q)    \\
  bcast(e, \_)          &= \bot \2 (mem,emitInt,break,if,  \\
                                 & \5\5 loop,and,or,@nop) %\\
\end{align*}
}
\caption{
The function $bcast$ awakes awaiting trails matching the event by converting
$awaitExt$ and $awaitInt$ to $@nop$ expressions, and by unwinding an $every$
expression.
\label{fig.bcast}
}
\end{figure}

\begin{figure}
{\small
\begin{align*}
  isblocked(n, \1 awaitExt(id)) &= true                                   \\
  isblocked(n, \1 awaitInt(id)) &= true                                   \\
  isblocked(n, \1 every~e~p)    &= true                                   \\
  isblocked(n, \1 @canrun(m))   &= (n > m)                                \\
  isblocked(n, \1 fin~p)        &= true                                   \\
  isblocked(n, \1 p~;~q)        &= isblocked(n,p)                         \\
  isblocked(n, \1 p~@loop~q)    &= isblocked(n,p)                         \\
  isblocked(n, \1 p~@and~q)     &= isblocked(n,p) \wedge isblocked(n,q)   \\
  isblocked(n, \1 p~@or~q)      &= isblocked(n,p) \wedge isblocked(n,q)   \\
  isblocked(n, \1 \_)           &= false \2 (mem,emitInt,break,if,        \\
                                & \5\5\5\1 loop,and,or,@nop)   %\\
\end{align*}
}
\caption{
The predicate $isblocked$ is true only if all branches in parallel are blocked
waiting for events, finalization clauses, or certain stack levels.
\label{fig.isblocked}
}
\end{figure}

\begin{figure}[b]
{\small
\begin{align*}
  clear( awaitExt(e) ) &= @nop                  \\
  clear( awaitInt(e) ) &= @nop                  \\
  clear( every~e~p )   &= @nop                  \\
  clear( @canrun(n) )  &= @nop                  \\
  clear( fin~p )       &= p                     \\
  clear( p~;~q )       &= clear(p)              \\
  clear( p~@loop~q )   &= clear(p)              \\
  clear( p~@and~q )    &= clear(p)~;~clear(q)   \\
  clear( p~@or~q )     &= clear(p)~;~clear(q)   \\
  clear( \_ )          &= \bot \2 (mem,emitInt,break,if, \\
                                  & \5\5 loop,and,or,@nop) %\\ 
\end{align*}
}
\caption{
The function $clear$ extracts $fin$ expressions in parallel and put their 
bodies in sequence.
\label{fig.formal.clear}
}
\end{figure}
